### 2022-03-08 symposium.tex
Related work WIP

Our presentation is structured as follows:

\begin{itemize}
  \item
\Cref{sec:background}
  \item
\Cref{sec:architecture},~\ref{sec:type-level} and~\ref{sec:term-level}
  \item
\Cref{sec:evaluation}
  \item
\Cref{sec:related}
\Cref{sec:conclusion}

\end{itemize}

@@

\subection{Match Types}

In this section, we give a brief introduction to match types by inspecting the following example:

\elemExample
% type Elem[X] = X match
%   case String => Char
%   case Iterable[t] => Elem[t]
%   case Any => X

\noindent
This example defines a type |Elem| parametrized by one type parameter |X|.
The right-hand side is defined in terms of a match on that type parameter---a match type.
A match type reduces to one of its right-hand sides, depending on the type of its scrutinee.
For example, the above type reduces as follows:

\begin{lstlisting}
Elem[String] ~> Char
Elem[Int] ~> Int
Elem[Iterable[Int]] ~> Int
\end{lstlisting}

@@

\subection{The Generic View of Tuples}

% https://www.scala-lang.org/2021/02/26/tuples-bring-generic-programming-to-scala-3.html

@@

We present our library in terms of 3 components:

@@

\noindent
The |Length| and |CharAt| types (from |ops.string|) are special-cased by the compiler: when their first argument is a known string literal, the compiler evaluates those types via their corresponding term-level implementation.
Similarly, the |+| type (from |ops.int|) allows us to manipulate integers just like we would at the term level.
The |EmptyTuple| and |*:| types are Scala~3's new generic representation of tuples.
Those types remove the need for heterogeneous lists and let us manipulate tuples directly.

@@

@@

In this \thesisOnly{chapter}\paperOnly{paper}, we introduced a new design for type-safe regular expressions in Scala.
We presented our type-level analysis, which identifies capturing groups and computes their nullability.
We built our library following the design of Scala's original regular expression API, in order to provide a frictionless migration path for programmers interested in the additional type safety.
We evaluated our implementation by running it against the QT3TS test suite and showed that, on those benchmarks, our type-level analysis only has a marginal impact on compilation times.

>>>>

This paper is structured as follows.
In \Cref{sec:background}, we give an introduction to match types and generic tuples, two recent additions to Scala's type system which we rely on in our implementation.
In \Cref{sec:architecture},~\ref{sec:type-level} and~\ref{sec:term-level}, we present our library for type-safe regular expressions.
In \Cref{sec:evaluation}, we evaluate the performance of our implementation in terms of compilation times and execution times.
In \Cref{sec:related}, we discuss related work.
We conclude the paper in \Cref{sec:conclusion}.

@@

In this section, we give a brief introduction to two recent additions to Scala's type system: match types and generic tuples.
Our regular expression library makes extensive use those two features to enable type-safe and null-safe capturing group extraction.

\subsection{Match Types}

Match types provide first-class support for type-level computations by allowing programmers to perform pattern matching on types:

\elemExample
% type Elem[X] = X match
%   case String => Char
%   case Array[t] => Elem[t]
%   case Any => X

\noindent
This example defines a type |Elem| parametrized by one type parameter |X|.
The right-hand side is defined in terms of a match on that type parameter -- a match type.
A match type reduces to one of its right-hand sides, depending on the type of its scrutinee.
For example, the above type reduces as follows:

\begin{lstlisting}
Elem[String] =:= Char
Elem[Int] =:= Int
Elem[Array[Int]] =:= Int
\end{lstlisting}

@@

\subsection{Generic Tuples}
\label{subsec:generic-tuples}

Scala's tuples were originally defined as plain old data types.
In Scala~3, tuples got enhanced with a generic representation~\citep{bazzucchi2021tuples}, similar to heterogeneous lists~\citep{kiselyov2004strongly}.
This representation uses two types, |EmptyTuple| and |*:|, to describe tuples in a list-like fashion.
The compiler treats those new types and the traditional class-based representation of tuples interchangeably.
A 2-tuple of integers, for example, has two equivalent representations:

\begin{lstlisting}
(Int, Int) =:= Int *: Int *: EmptyTuple
\end{lstlisting}

\noindent
This new tuple representation is especially useful at the type level, where it allows programmers to manipulate tuples in a recursive manner.
For example, the following match type reverses the order of a tuple's elements:

\tupleReverseA
\vspace{-3pt}
\tupleReverseB

@@

We present our library for type-safe regular expressions in terms of 3 components:

@@

\noindent
The |Length| and |CharAt| types (from |ops.string|) are special-cased by the compiler: when their first argument is a known string literal, the compiler evaluates those types via their corresponding term-level implementation.
Similarly, the |+| type (from |ops.int|) allows us to manipulate integers just like we would at the term level.
The |EmptyTuple| and |*:| types are Scala~3's new generic representation of tuples, presented in \Cref{subsec:generic-tuples}.

@@

The typed-regex~\citep{akshay2021typed} Typescript library provides a type-safe API for extracting named capturing groups of regular expressions.
The implementation uses conditional types~\citep{microsoft2020typescript}, Typescript's type-level ternary operator, to statically analyze regular expressions.
Their approach, like ours, identifies optional capturing groups and marks them as such (using Typescript's optional properties).
At the time of writing, the library is still in an early stages of development and only supports a subset of the regular expression syntax, and notably lacks support for nested optional capturing group detection.

The safe-regex-rs~\citep{leonhard2021safe} Rust library takes a different approach towards the same goal: it uses Rust macros to code generate arity-specific extractors.
Unlike the
Their implementation is

Rust \url{https://gitlab.com/leonhard-llc/safe-regex-rs}

@@

In this \thesisOnly{chapter}\paperOnly{paper}, we introduce a new design for type-safe regular expressions in Scala.
We presented our type-level analysis, which identifies capturing groups and computes their nullability.
We built our library following the design of Scala's original regular expression API, in order to provide a frictionless migration path for programmers interested in the additional type safety.
We evaluated our implementation by running it against the QT3TS test suite and showed that, on those benchmarks, our type-level analysis only has a marginal impact on compilation times.
